---
editor_options:
  chunk_output_type: console
---

# INTERNAL

## Package load/unload

```{r}
.onLoad <- function(libname, pkgname) {

  # cf. https://rstudio.github.io/reticulate/articles/python_dependencies.html#onload-configuration
  reticulate::configure_environment(package = pkgname)
}
```

## Avoid `R CMD check` notes about undefined global objects used in magrittr pipes

cf. <https://github.com/tidyverse/magrittr/issues/29#issuecomment-74313262>

```{r}
utils::globalVariables(names = c(".",
                                 "padding_bottom",
                                 "padding_left",
                                 "padding_right",
                                 "padding_top",
                                 "show_progress"))
```

## Constants

```{r}
this_pkg <- utils::packageName()
```

### `formats_postscript`

Based on what SVG images can be converted to by the different `rsvg::rsvg_*()` functions.

```{r}
formats_postscript <- c("pdf", "eps", "ps")
```

### `formats_raster`

Based on what SVG images can be converted to by the different `rsvg::rsvg_*()` functions.

```{r}
formats_raster <- c("png", "webp")
```

## Functions

### `post_process_svg`

```{r}
post_process_svg <- function(path,
                             responsive,
                             crop,
                             padding) {
  
  checkmate::assert_file_exists(path,
                                access = "w")
  checkmate::assert_flag(responsive)
  checkmate::assert_flag(crop)
  checkmate::assert_numeric(padding,
                            lower = 0.0,
                            any.missing = FALSE,
                            min.len = 1L,
                            max.len = 4L)
  
  xml <- xml2::read_xml(path)
  attrs <- character()
  
  # change SVGs' `width` and `height` properties to `100%` if requested
  if (responsive) {
    attrs["width"] <- "100%"
    attrs["height"] <- "100%"
  }
  
  # crop unused transparent space around image if requested; inspired by https://stackoverflow.com/a/69783110/7196903
  if (crop) {
    
    # init padding
    n_padding <- length(padding)
    padding_top <- padding_bottom <- padding_left <- padding_right <- padding[1L]
    
    if (n_padding == 4L) {
      padding_right <- padding[2L]
      padding_bottom <- padding[3L]
      padding_left <- padding[4L]
    } else if (n_padding == 3L) {
      padding_right <- padding[2L]
      padding_left <- padding[2L]
      padding_bottom <- padding[3L]
    } else if (n_padding == 2L) {
      padding_right <- padding[2L]
      padding_left <- padding[2L]
    }
    
    # determine opaque pixel coordinates
    data_img <- magick::image_read_svg(path = path) |> magick::image_data()
    ix_opaque <- which(data_img[4L, , ] != 0L,
                       arr.ind = TRUE)
    ix_opaque_start <- apply(X = ix_opaque,
                             MARGIN = 2L,
                             FUN = min)
    ix_opaque_end <- apply(X = ix_opaque,
                           MARGIN = 2L,
                           FUN = max) - ix_opaque_start
    
    attrs["viewBox"] <- paste(paste(ix_opaque_start - c(padding_left, padding_top),
                                    collapse = " "),
                              paste(ix_opaque_end + c(padding_left + padding_right, padding_top + padding_bottom),
                                    collapse = " "))
  }
  
  # apply new SVG attributes
  # NOTE: we can't use `xml2::xml_set_attrs()` since it completely replaces all attrs but at the same time doesn't allow to set the
  #        `xmlns:xlink` attr again (it fails with a critical error if it's tried); thus we set the additional attrs one by one
  purrr::iwalk(attrs,
               \(val, name) xml2::xml_set_attr(x = xml,
                                               attr = name,
                                               value = val))
  # overwrite original SVG file
  xml2::write_xml(x = xml,
                  file = path,
                  options = c("format", "no_declaration"))
  
  invisible(path)
}
```

### `svg_to_postscript`

```{r}
svg_to_postscript <- function(path,
                              format = formats_postscript,
                              crop,
                              padding) {
  
  checkmate::assert_file_exists(path,
                                access = "r")
  format <- rlang::arg_match(format)
  checkmate::assert_numeric(padding,
                            lower = 0.0,
                            any.missing = FALSE,
                            min.len = 1L,
                            max.len = 4L)
  checkmate::assert_flag(crop)
  path_output <- fs::path_ext_set(path,
                                  ext = format)
  switch(format,
         eps = rsvg::rsvg_eps(svg = path,
                              file = path_output),
         pdf = rsvg::rsvg_pdf(svg = path,
                              file = path_output),
         ps = rsvg::rsvg_ps(svg = path,
                            file = path_output))
  
  # cropping is only supported for PDF
  if (format == "pdf" && crop) {
    
    pal::assert_cli(
      cmd = "pdfcrop",
      error_msg = paste0("The {.strong pdfcrop} executable is required but couldn't be found on system's {.href ",
                         "[PATH](https://en.wikipedia.org/wiki/PATH_(variable))}. It is usually installed together with a full ",
                         "TeX distribution like {.href [TeX Live](https://en.wikipedia.org/wiki/TeX_Live)}. If you're using ",
                         "{.href [TinyTeX](https://yihui.org/tinytex/)}, simply install the {.href [pdfcrop LaTeX ",
                         "package](https://ctan.org/pkg/pdfcrop)} via {.run tinytex::tlmgr_install(\"pdfcrop\")}.")
    )
    
    if (tools::find_gs_cmd() == '') {
      cli::cli_alert_warning(paste0("PDF cropping is skipped since the required Ghostscript executable wasn't found. See",
                                    "{.fun tools::find_gs_cmd} for how it is detected on the system."))
    } else {
      
      # init padding
      n_padding <- length(padding)
      padding_top <- padding_bottom <- padding_left <- padding_right <- padding[1L]
      
      if (n_padding == 4L) {
        padding_right <- padding[2L]
        padding_bottom <- padding[3L]
        padding_left <- padding[4L]
      } else if (n_padding == 3L) {
        padding_right <- padding[2L]
        padding_left <- padding[2L]
        padding_bottom <- padding[3L]
      } else if (n_padding == 2L) {
        padding_right <- padding[2L]
        padding_left <- padding[2L]
      }
      
      system2(command = "pdfcrop",
              args = c(glue::glue("--margins '{padding_left} {padding_top} {padding_right} {padding_bottom}'"),
                       path_output,
                       path_output),
              stdout = FALSE)
    }
  }
}
```

### `svg_to_raster`

Raster image formats do not properly handle viewbox-cropped SVGs when converted via rsvg.

```{r}
svg_to_raster <- function(path,
                          format = formats_raster) {
  
  checkmate::assert_file_exists(path,
                                access = "r")
  format <- rlang::arg_match(format)
  path_output <- fs::path_ext_set(path,
                                  ext = format)
  switch(format,
         png = rsvg::rsvg_png(svg = path,
                              file = path_output),
         webp = rsvg::rsvg_webp(svg = path,
                                file = path_output))
}
```

### `write_svg`

```{r}
write_svg <- function(plots,
                      dir,
                      width,
                      height,
                      scale,
                      show_progress,
                      msg) {
  
  # ensure required Python modules are available
  has_python <-
    c("plotly", "kaleido") %>%
    purrr::map_lgl(reticulate::py_module_available) %>%
    all()
  
  if (!has_python) {
    cli::cli_abort(paste0("A working Python setup including the modules {.field plotly} and {.field kaleido} is required. See ",
                          "{.url https://rdrr.io/cran/plotly/man/save_image.html#heading-4} for details."))
  }
  
  # initialize kaleido
  kaleido <- plotly::kaleido()
  
  # disable MathJax to avoid "Loading [MathJax]/extensions/MathMenu.js" message
  # TODO: remove this once https://github.com/plotly/Kaleido/issues/122 is resolved
  kaleido$scope$mathjax <- NULL
  
  # export to SVG
  purrr::pwalk(.l = purrr::compact(list(obj = plots,
                                        id = names(plots),
                                        width = width,
                                        height = height,
                                        scale = scale)),
               .f = \(obj,
                      id,
                      width = NULL,
                      height = NULL,
                      scale = NULL) {
                 kaleido$transform(p = obj,
                                   file = fs::path(dir, id,
                                                   ext = "svg"),
                                   width = width,
                                   height = height,
                                   scale = scale)
               },
               .progress = ifelse(show_progress,
                                  msg,
                                  FALSE))
  kaleido$shutdown()
}
```

# EXPORTED

## `write_img`

NOTES:

-   To initialize reticulate's default shared Python virtualenv `r-reticulate`, run `reticulate::virtualenv_create()`. Before reticulate
    [v1.31](https://github.com/rstudio/reticulate/releases/tag/v1.31), it nudged to install miniconda if this virtualenv didn't exist. Latest reticulate
    automatically proposes to create this virtualenv 🎉!

-   `rsvg::rsvg_*()` handles unused transparent space around the input SVG differently for the different output formats. `rsvg::rsvg_eps/pdf/ps()` do not
    properly handle viewbox-cropped SVGs, while `rsvg::rsvg_png/webp` do. Thus we account for this in the function's logic.

-   The resulting SVG images are not fully reproducible/deterministic due to SVG element `id` attributes including a random hash value (e.g. `defs-cf4626`,
    `topdefs-cf4626`, `legendcf4626` etc.). Setting RNG seeds, i.e. `set.seed(42L); htmlwidgets::setWidgetIdSeed(seed = 42)`, before the Kaleido export doesn't
    help -- which means this probably needs to rather be addressed in the underlying Kaleido Python code instead.

```{r}
#' Save Plotly charts as static images
#'
#' @description
#' Writes Plotly charts to static image files in SVG and optionally additional formats (EPS, PDF, PNG, PS, and/or WebP) using [plotly::kaleido()].
#' 
#' In contrast to [plotly::save_image()], this function
#'
#' - can export multiple [plotly objects][plotly::plot_ly] in a single kaleido session to multiple output formats in one go.
#' - can tweak the exported SVG images to be responsive.
#' - supports automatic cropping of unused space around the exported images.
#'
#' @details
#' PNG and WebP images are completely based on the resulting SVG image, while EPS, PDF and PS are built from an intermediate SVG representation (before
#' viewbox-based SVG cropping is applied). Hence, results of the different output formats might differ.
#'
#' It's recommended to rely on SVG images wherever possible and fall back to PDF images where necessary (e.g. in LaTeX documents) since these two vector image
#' formats provide the best visual results. EPS and PS images should be avoided if possible due to their shortcomings in cropping and padding.
#'
#' @param plots Charts to export. A named list of [plotly objects][plotly::plot_ly]. Names are used as output filenames (excl. filetype extension).
#' @param dir Path to the directory of the exported files. A character scalar.
#' @param formats Additional image file formats to export to besides the default SVG format. Zero or more of
#'   `r c(formats_raster, formats_postscript) |> pal::as_md_vals() |> pal::prose_ls()`.
#' @param width Width of the exported image in layout pixels. If `scale` is 1, this will also be the width of the exported image in physical pixels. An integer
#'   vector that is recycled to the length of `plots`, or `NULL`.
#' @param height Height of the exported image in layout pixels. If `scale` is 1, this will also be the height of the exported image in physical pixels. An
#'   integer vector that is recycled to the length of `plots`, or `NULL`.
#' @param scale Scale factor to use when exporting the figure. A scale factor larger than `1.0` will increase the image resolution with respect to the figure's
#'   layout pixel dimensions. Whereas as scale factor of less than 1.0 will decrease the image resolution. A numeric vector that is recycled to the length of
#'   `plots`, or `NULL`.
#' @param responsive Whether or not to modify the SVG image to become responsive when inlined into HTML by setting its `width` and `height` property to
#'   `"100%"`.A logical vector, recycled to the length of `plots`.
#' @param crop Whether or not to crop unused space around the generated image. Note that this has no effect on `formats = c("eps", "ps")` – EPS images are
#'   always cropped while cropping PS images is not supported. A logical vector, recycled to the length of `plots`.
#' @param padding Padding to leave after cropping unused space around the generated image. Either a single numeric vector, or a list of numeric vectors. Each
#'   vector must be of length 1–4:
#'   
#'   - When one value is specified, it applies the same padding to all four sides.
#'   - When two values are specified, the first padding applies to the top and bottom, the second to the left and right.
#'   - When three values are specified, the first padding applies to the top, the second to the right and left, the third to the bottom.
#'   - When four values are specified, the paddings apply to the top, right, bottom, and left in that order (clockwise).
#'   
#'   For `formats = "pdf"`, the values are in [`bp`](https://en.wikibooks.org/wiki/LaTeX/Lengths#Units) ("big points"), for all other formats in pixels. Only
#'   relevant if `crop = TRUE`. Note that `padding` has no effect on `formats = "eps"`, i.e. EPS images are always fully cropped.
#' @param show_progress `r pkgsnip::param_lbl("show_progress")`
#'
#' @return `plots`, invisibly.
#' @export
#'
#' @examples
#' \dontrun{
#' plotly::plot_ly(data = mtcars,
#'                 type = "scatter",
#'                 mode = "markers",
#'                 x = ~mpg,
#'                 y = ~hp) |>
#'   list("mtcars_mpg_by_hp" = _) |>
#'   plotlee::write_img(formats = c("pdf", "png"))}
write_img <- function(plots,
                      dir = ".",
                      formats = "pdf",
                      width = NULL,
                      height = NULL,
                      scale = NULL,
                      responsive = TRUE,
                      crop = TRUE,
                      padding = c(0L, 10L, 0L, 10L),
                      show_progress = TRUE) {
  
  checkmate::assert_list(plots,
                         types = "plotly",
                         any.missing = FALSE,
                         names = "unique")
  n_plots <- length(plots)
  checkmate::assert_directory_exists(dir,
                                     access = "w")
  checkmate::assert_subset(formats,
                           choices = c(formats_raster, formats_postscript))
  checkmate::assert_integer(width,
                            lower = 0L,
                            any.missing = FALSE,
                            null.ok = TRUE)
  checkmate::assert_integer(height,
                            lower = 0L,
                            any.missing = FALSE,
                            null.ok = TRUE)
  checkmate::assert_numeric(scale,
                            lower = 0.0,
                            any.missing = FALSE,
                            null.ok = TRUE)
  checkmate::assert_logical(responsive,
                            any.missing = FALSE)
  checkmate::assert_logical(crop,
                            any.missing = FALSE)
  
  if (is.list(padding)) {
    checkmate::assert_list(padding,
                           any.missing = FALSE,
                           len = n_plots)
  } else {
    checkmate::assert_numeric(padding,
                              lower = 0.0,
                              any.missing = FALSE,
                              min.len = 1L,
                              max.len = 4L)
  }
  checkmate::assert_flag(show_progress)
  
  # recycle args to common length
  if (!is.null(width)) width %<>% vctrs::vec_recycle(size = n_plots,
                                                     x_arg = "width")
  if (!is.null(height)) height %<>% vctrs::vec_recycle(size = n_plots,
                                                       x_arg = "height")
  if (!is.null(scale)) scale %<>% vctrs::vec_recycle(size = n_plots,
                                                     x_arg = "scale")
  responsive %<>% vctrs::vec_recycle(size = n_plots,
                                     x_arg = "responsive")
  crop %<>% vctrs::vec_recycle(size = n_plots,
                               x_arg = "crop")
  # normalize dir
  dir %<>% fs::path_real()
  dir_rel <-
    dir |>
    c(fs::path_wd()) |>
    fs::path_common() |>
    fs::path_rel(path = dir)
  
  # export to SVG
  write_svg(plots = plots,
            dir = dir,
            width = width,
            height = height,
            scale = scale,
            show_progress = show_progress,
            msg = cli::format_inline("Exporting {.val {length(plots)}} Plotly chart{?s} as static {.field SVG} and {.field {toupper(formats)}}",
                                     "{cli::qty(length(plots))} image{?s} to {.path {paste0(dir_rel, '/')}}..."))
  
  # convert SVGs to those additional requested formats that **don't** properly handle viewbox-cropped SVGs
  all_svg_paths <- fs::path_ext_set(path = fs::path(dir, names(plots)),
                                    ext = "svg")
  purrr::walk(formats_postscript,
              \(format) {
                
                purrr::pwalk(.l = list(path = all_svg_paths,
                                       format = format,
                                       crop = crop,
                                       padding = padding),
                             .f = svg_to_postscript)
              },
              .progress = ifelse(show_progress,
                                 cli::format_inline(paste0("Converting {.val {length(all_svg_paths)}} original SVG images to {.field ",
                                                           "{formats_postscript}} formats")),
                                 FALSE))
  
  # post-process SVGs if requested
  i_postprocess <- which(responsive | crop)
  
  purrr::pwalk(.l = list(path = all_svg_paths[i_postprocess],
                         responsive = responsive[i_postprocess],
                         crop = crop[i_postprocess],
                         padding = padding[i_postprocess]),
               .f = post_process_svg,
               .progress = ifelse(show_progress,
                                  cli::format_inline("Post-processing {.val {length(plots)}} SVG images"),
                                  FALSE))
  
  # convert SVGs to the remaining additional requested formats that **do** properly handle viewbox-cropped SVGs
  purrr::walk(formats_raster,
              \(format) purrr::pwalk(.l = list(path = all_svg_paths,
                                               format = format),
                                     .f = svg_to_raster),
              .progress = ifelse(show_progress,
                                 cli::format_inline(paste0("Converting {.val {length(all_svg_paths)}} post-processed SVG images to ",
                                                           "{.field {formats_raster}} formats")),
                                 FALSE))
  invisible(plots)
}
```

## `simplify_trace_ids`

NOTES:

-   The underlying issue this function solves is reported upstream in [#2074](https://github.com/plotly/plotly.R/issues/2074). See also [this
    comment](https://github.com/plotly/plotly.R/pull/1365#issuecomment-434063157).

```{r}
#' Make plotly trace identifiers reproducible
#'
#' Replaces the random trace identifiers of a [plotly object][plotly::plot_ly] with simple sequential ones (`trace_1`, `trace_2` and so on).
#'
#' This function is especially useful to apply before a plotly object or its JSON representation (as generated for rendered R Markdown / Quarto documents) 
#' is [version controlled](https://en.wikipedia.org/wiki/Version_control) (e.g. via Git).
#' 
#' To also make the htmlwidget identifier reproducible when rendering a Plotly chart, use [htmlwidgets::setWidgetIdSeed()].
#'
#' @param p [Plotly object][plotly::plot_ly] to modify.
#'
#' @return `r pkgsnip::return_lbl("plotly_obj")`
#' @export
#'
#' @examples
#' p <- plotly::plot_ly(data = mtcars,
#'                      x = ~mpg,
#'                      type = "histogram")
#'
#' p2 <- plotly::plot_ly(data = mtcars,
#'                       x = ~mpg,
#'                       type = "histogram")
#'
#' p_deterministic <- plotlee::simplify_trace_ids(p)
#' p2_deterministic <- plotlee::simplify_trace_ids(p2)
#'
#' names(p$x$visdat)
#' names(p2$x$visdat)
#' names(p_deterministic$x$visdat)
#' names(p2_deterministic$x$visdat)
simplify_trace_ids <- function(p) {
  
  trace_ids <- names(p$x$visdat)
  cur_id <- p$x$cur_data
  
  for (i in seq_along(trace_ids)) {
    
    new_id <- paste0("trace_", i - 1L)
    
    p$x %<>% purrr::modify_tree(is_node = is.list,
                                post = \(x) {
                                  
                                  if (utils::hasName(x, trace_ids[i])) {
                                    names(x)[names(x) == trace_ids[i]] <- new_id
                                  }
                                  
                                  x
                                })
    
    if (trace_ids[i] == cur_id) {
      p$x$cur_data <- new_id
    }
  }
  
  p
}
```
